Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-11-08 intro.tex
rewrite intro

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Introduction}
\label{chapter:intro}
The standard model of particle physics (SM) is our current best working
theory that describes the phenomenon we observe, for  example,
 at collider experiments.
The large hadron collider (LHC) is a high energy proton-proton
collider experiment that allows us to study the fundamental particles
laid out in the standard model.

The interaction of these fundamental particles are described
by scattering amplitudes. The standard model allows us to
calculate these amplitudes, or matrix elements, from first
principles. A key challenge in the past few decades has been
to compute these matrix elements for increasing number of
final state particles and for higher orders in perturbative
expansions in fixed order calculations. Although there
has been tremendous progress in this sector a complication
that naturally arises is that the evaluation of these matrix
elements becomes increasingly time-consuming due to their
algebraic complexity.

The topic of this thesis is to apply modern machine learning
techniques to reduce the time spent evaluating these
complex matrix elements. Considering the physical properties
of these matrix elements allows us to implement physics
knowledge into these machine learning algorithms giving
us higher levels of performance than would otherwise be
achievable.

The structure of this thesis is as follows: in Chapter~\ref{chapter:intro}
I will recap the necessary concepts from the standard
model, in particular, on quantum chromodynamics (QCD), the sector
which governs the behaviour of quarks and gluons. Furthermore,
the relationship between theory and experiment will be elaborated on.

In Chapter~\ref{chapter:qcd} I will discuss the challenges that arise
in fixed order perturbative calculations and the current methods
that have been adopted in the community to deal with them. These
methods also form the basis of our physics knowledge which we
embed into the machine learning algorithm. This procedure is 
described in detail in Chapter~\ref{chapter:fame1} for electron-positron
annihilation for tree-level matrix element emulation.
A similar philosophy is applied in Chapter~\ref{chapter:fame2}
where next-to-leading order QCD k-factors are emulated.
These two chapters discuss in detail the procedure for
electron-positron annihilation, however, the most relevant
experiment currently is the LHC which is colliding protons.
The extension to hadron-hadron colliders is detailed in
Chapter~\ref{chapter:fame3} where we explore using the
emulator in a novel implementation to replace traditional
matrix element providers in the event generator SHERPA.

Finally, I will conclude the thesis in Chapter~\ref{chapter:conclusion}.

\section{The Standard Model of particle physics}
    The Standard Model of particle physics (SM) is our current
    best working theory to describe all known elementary particles
    as well as three of the four fundamental forces. Developed
    predominantly in the latter half of the 20th century, it is
    one of the most well tested theories we have in science today.
    Some highlights include the highly precise predictions of the
    anomalous magnetic moment of the electron, agreeing with experimental
    measurements to more than 10 significant figures \cite{Aoyama:2017uqe},
    and the discovery of the Higgs boson in 2012 by the ATLAS \cite{ATLAS:2012yve}
    and CMS experiments \cite{CMS:2012qbp} at the Large Hadron Collider
    (LHC) which was theorised decades prior.

@@

    The gauge coupling of the group $g_{s}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{s} = \dfrac{g_{s}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process. This
    is due to the process of renormalisation which will be discussed
    in Section~\ref{sec:renormalisation}.
    The implication of this is that at collider
    experiments where collisions occur at extremely high energy,
    the strong coupling constant becomes small, a
    property known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{s}$.

>>>>

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Introduction}
\label{chapter:intro}

The Large Hadron Collider (LHC) at CERN has been colliding
particles since 2009 with each successive run increasing the centre-of-mass
energy of the collision (7-13.6 TeV). With the experiment operating
over the course of decades, there is a vast quantity of
experimental data collected that has to be compared with
theoretical predictions. Currently, the theory predictions
are made with the Standard Model (SM) of
particle physics which describes the fundamental particles
of nature and their interactions.

The standard paradigm for comparing
theory and experiment is to simulate the particle collisions
from the initial collision all the way through
to the detection of the hundreds of particles produced in the collision.
General purpose Monte Carlo event generators are the de-facto
tool designed to generate these simulated events.
With the large number of scattering events collected at
the LHC already, and with even more expected at the High Luminosity LHC,
the rate of event generation is at risk of
being outpaced by experimental data collection.

The process of generating simulated events can be broadly
split up into three segments:
the highly energetic hard scattering process, 
the subsequent cascading of particles (parton shower),
and the formation of bound states detected (hadronisation).
While hadronisation is based on phenomological models,
the hard scattering and parton shower are derived
from the SM and are theoretically well-motivated.

The hard scattering process can be formulated as a probability
for particles to collide, represented by matrix elements.
These matrix elements are formally calculated order-by-order
in perturbation theory with each order leading
to increasingly difficult computations. In practice,
these matrix elements are provided in libraries interfaced
to the event generators and evaluation is largely automated for many
of the most important processes. However, the evaluation time
for these complex expressions becomes problematic
when generating the large samples required to compare with
experiments, especially when a significant portion of time
spent simulating an event is spent on the matrix element
evaluation.

The focus of this thesis will be on applying modern
machine learning methods to accelerate the evaluation
of matrix elements, hence speeding up the event
generation process in general. Machine learning algorithms
have become ubiquitous in many fields and their applicability
to a wide range of problems have made them a popular choice
in the high energy physics community as well. By combining the
well understood behaviour of matrix elements in specific
kinematic regions with powerful machine learning
algorithms, the construction of physics-inspired machine
learning models unlocks higher levels of performance
than would otherwise be achievable.

The structure of this thesis is as follows: in this chapter
I will recap the necessary concepts from the SM,
in particular quantum chromodynamics (QCD), the sector
which governs the behaviour of quarks and gluons. Furthermore,
the relationship between theory and experiment will be elaborated on
with the introduction of cross-sections and their relation
to matrix elements.

In Chapter~\ref{chapter:qcd}, I will discuss the challenges that arise
in fixed-order perturbative calculations and the current methods
that have been adopted in the community to deal with them. Some of
these methods also form the theoretical foundation which we
embed into the machine learning algorithm. This procedure is 
described in detail in Chapter~\ref{chapter:fame1} for electron-positron
annihilation for tree-level matrix element emulation.
A similar philosophy is applied in Chapter~\ref{chapter:fame2}
where next-to-leading order QCD k-factors are emulated for
the same processes.
These two chapters discuss in detail the procedure for
constructing emulators for electron-positron annihilation,
however, the more relevant processes for the LHC are in
proton-proton initiated collisions. The extension
to this scenario is detailed in Chapter~\ref{chapter:fame3}
where we also explore using the
emulator in a novel implementation to replace traditional
matrix element providers in the event generator SHERPA.
Finally, I will summarise and conclude the thesis in
Chapter~\ref{chapter:conclusion}.

\section{The Standard Model of particle physics}
    The Standard Model of particle physics (SM) is our current
    best working theory to describe all known elementary particles
    as well as three of the four fundamental forces. Developed
    predominantly in the latter half of the 20th century, it is
    one of the most well tested theories we have in science today.
    Some highlights include the highly precise predictions of the
    anomalous magnetic moment of the electron, agreeing with experimental
    measurements to more than 10 significant figures \cite{Aoyama:2017uqe},
    and the discovery of the Higgs boson in 2012 by the ATLAS \cite{ATLAS:2012yve}
    and CMS experiments \cite{CMS:2012qbp} at the LHC
    which was theorised decades prior.

@@

    The gauge coupling of the group $g_{s}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{s} = \dfrac{g_{s}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process. This
    is due to the process of renormalisation which will be discussed
    in Section~\ref{sec:renormalisation}.
    The implication of this is that at collider
    experiments where collisions occur at extremely high energy,
    the strong coupling constant becomes small, a
    property known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{s}$ (see Section~\ref{sec:xs}).

### 2022-11-08 qcd.tex
fix typos
rewrite intro

    In an all orders calculation of an observable,
    there would be no dependence on the renormalisation
    scale as it is a remnant of the regularisation prescription
    However, because in perturbative QCD observables
    are calculated order-by-order, there will be a residual
    dependence on the renormalisation scale stemming from
    the missing higher order terms. The conventional way
    that the community estimates the uncertainty associated
    with this residual dependence is by carrying out
    a scale variation (see Section~\ref{sec:sec:scale_variations}).

@@

    Below we list the antenna functions relevant
    for multijet production in electron-positron
    annihiliation with a single quark flavour,
    namely the partonic channel $e^{-}e^{+} \rightarrow
    q \bar{q}$ + $n \, g$. The emulation of the NLO QCD K-factors
    of this partonic process is examined in Chapter~\ref{chap:fame2}.
    For this process we require the $A$, $D$, and $F$ antenna functions,
    which are illustrated in Figure~\ref{fig:antenna}.
    The $i,j,k$ indices will be fixed such that $q = 1$, $\bar{q} = 2$,
    and $g \in \{3, 4, 5\}$ in view of using the
    antenna functions where the momenta
    of partons will have fixed indices in a phase-space
    point represented as an array.

>>>>

    In an all orders calculation of an observable,
    there would be no dependence on the renormalisation
    scale as it is a remnant of the regularisation prescription
    However, because in perturbative QCD observables
    are calculated order-by-order, there will be a residual
    dependence on the renormalisation scale stemming from
    the missing higher order terms. The conventional way
    that the community estimates the uncertainty associated
    with this residual dependence is by carrying out
    a scale variation (see Section~\ref{sec:scale_variations}).

@@

    Below we list the antenna functions relevant
    for multijet production in electron-positron
    annihiliation with a single quark flavour,
    namely the partonic channel $e^{-}e^{+} \rightarrow
    q \bar{q}$ + $n \, g$. The emulation of the NLO QCD K-factors
    of this partonic process is examined in Chapter~\ref{chapter:fame2}.
    For this process we require the $A$, $D$, and $F$ antenna functions,
    which are illustrated in Figure~\ref{fig:antenna}.
    The $i,j,k$ indices will be fixed such that $q = 1$, $\bar{q} = 2$,
    and $g \in \{3, 4, 5\}$ in view of using the
    antenna functions where the momenta
    of partons will have fixed indices in a phase-space
    point represented as an array.
